Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-09 paper.qmd
finished with Section 5, Curiosities

Differences in density can arise by sampling at different rates in different subspaces of \gD{}. For example, the data shown in @fig-one-dens_clust-error all lies on a \2D{} curved sheet in \fD{}, but one end of the sheet is sampled densely and the other very sparsely. It was simulated to illustrate the effect of the density difference on layout generated by an NLDR, illustrated using the tSNE results.

@fig-one-dens_clust-error (a2, b2) shows \gD{} layout for tSNE created using the default hyper-parameters. One would expect to see a rectangular shape if the curved sheet is flattened, but the layout is triangular. The other two displays show the residuals as a dot density plot (a1, b1), and a \2D{} projection of the data and the model from \fD{} (a3, b3). Using the linked brushing tool, we can highlight points in the residuals

The \gD{} layout is colored according to the \fD{} model error (@fig-one-dens_clust-error a). In @fig-one-dens_clust-error a, the large model error points are clustered in the top corner, forming a separate cluster with high model errors.

The model is fitted to this layout. @fig-one-dens_clust-error c shows the \gD{} projection of the fitted model. In this projection, the distribution of \fD{} model errors can be observed. The dense C-shaped structure shows high model errors at the sparse end. In the dense C-shaped structure, the high model errors occur because there are fewer data points positioned within the bins at the sparse end of the structure.

>>>>

Differences in density can arise by sampling at different rates in different subspaces of \pD{}. For example, the data shown in @fig-one-dens_clust-error all lies on a \2D{} curved sheet in \fD{}, but one end of the sheet is sampled densely and the other very sparsely. It was simulated to illustrate the effect of the density difference on layout generated by an NLDR, illustrated using the tSNE results.

@fig-one-dens_clust-error (a2, b2) shows \gD{} layout for tSNE created using the default hyper-parameters. One would expect to see a rectangular shape if the curved sheet is flattened, but the layout is triangular. The other two displays show the residuals as a dot density plot (a1, b1), and a \gD{} projection of the data and the model from \fD{} (a3, b3). Using linked brushing between the plots, we can highlight points in the tSNE layout, and examine where they fall in the original \fD{}. The darker (maroon) points indicate points that have been highlighted by linking. In row a, the points at the top of the triangle are highlighted, and we can see these correspond to higher residuals, and also to all points at the low density end of the curved sheet. In row b, points at the lower left side of the triangle are highlighted which corresponds to smaller residuals and one corner of the sheet at the high density end of the curved sheet.

The tSNE behaviour is to squeeze the low density area of the data together into the layout. This is common in other NLDR methods also, which means analysts needs to be aware the if their data is not sampled relatively uniformly apparent closeness in the \gD{} may correspond to sparseness in \pD{}.
